---
title: "PPMI variables"
author: "Patrice Godard"
date: "`r format(Sys.time(), '%B %d %Y')`"
abstract: "This document describes how PPMI variables were derived from the original parsed data."
output: 
  BiocStyle::html_document: 
    fig_width: 9
    fig_height: 6.5
    toc: yes
editor_options: 
  chunk_output_type: console
---

<!-------------------------------------------------->
<!-------------------------------------------------->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls())
null <- gc()
try(null <- dev.off(), silent=T)
rm(null)
library(knitr)
library(beeswarm)
library(RColorBrewer)
load("PPMI-Parsed.rda")
varDoc$Source <- "Provided"
#######
beeswarm <- function(
    formula,
    data,
    corral="density",
    corralWidth=0.8,
    do.plot=TRUE,
    ...
){
    if(corral!="density"){
        toRet <- do.call(beeswarm::beeswarm, c(
            list(
                formula=formula,
                data=data,
                do.plot=F,
                corral=corral,
                corralWidth=corralWidth
            ),
            list(...)
        ))
    }else{
        toRet <- do.call(beeswarm::beeswarm, c(
            list(
                formula=formula,
                data=data,
                do.plot=F
            ),
            list(...)
        ))
        toRet <- do.call(rbind, by(
            toRet,
            toRet$x.orig,
            function(d){
                cpos <- round(mean(d$x, na.rm=T))
                xos <- d$x - cpos
                div <- max(abs(xos), na.rm=T)/(corralWidth/2)
                d$x <- cpos + (xos/div)
                return(d)
            }
        ))
    }
    if(do.plot){
        points(
            toRet$x, toRet$y,
            pch=toRet$pch,
            col=toRet$col, bg=toRet$bg
        )
    }
    invisible(toRet)
}
#######
rebox <- function(bp, boxwex=0.8, ...){
    sp <- boxwex/2
    for(i in 1:ncol(bp$stats)){
        y <- bp$stats[2:4, i]
        segments(
            x0=rep(i-sp, length(y)),
            x1=rep(i+sp, length(y)),
            y0=y,
            y1=y,
            ...
        )
    }
}
#######
mdSan <- function(txt){
    toRet <- txt
    toRet <- gsub('[_]', '\\\\_', toRet)
    return(toRet)
}
#######
vPatient <- function(d){
    sub("[.].*$", "", rownames(d))
}
vEvent <- function(d){
    sub("^.*[.]", "", rownames(d))
}
```


<!-------------------------------------------------->
<!-------------------------------------------------->

# Introduction

This document describes how PPMI variables were derived from the
original parsed data.

Derived variables were computed as described in the
<!-- *PPMI\_Derived\_Variable\_Definitions\_and\_Score\_Calculations20151201.pdf* --> <!-- For 2016-01-11 and before -->
*Derived\_Variable\_Definitions\_and\_Score\_Calculations.csv*
file.

<!-------------------------------------------------->
<!-------------------------------------------------->

# Demographics data

## Simple gender

The original provided gender
(`r paste(sort(unique(patientData$GENDER)), collapse=", ")`) were simplified to
*Female* or *Male* values.

```{r, echo=FALSE, message=FALSE}
patientData[,"SimpleGender"] <- as.factor(ifelse(
  patientData$GENDER=="Male",
  "Male",
  "Female"
))
parDocToAdd <- data.frame(
    Variable="SimpleGender",
    Description="Simplified gender: Male or Female",
    Group="Patient",
    "Sub-group"="Demographic",
    Class=class(patientData$SimpleGender),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
   )
rownames(parDocToAdd) <- "SimpleGender"
varDoc <- rbind(varDoc, parDocToAdd)
```

## Age

```{r, echo=FALSE, message=FALSE}
patientData[,"ENROLL_AGE"] <- as.double(
    patientData$ENROLL_DATE - patientData$BIRTHDT, units="days"
)/365.25
parDocToAdd <- data.frame(
    Variable="ENROLL_AGE",
    Description="Age at enrollment",
    Group="Patient",
    "Sub-group"="Demographic",
    Class=class(patientData$ENROLL_AGE),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "ENROLL_AGE"
varDoc <- rbind(varDoc, parDocToAdd)
```

## Origin population

```{r, echo=FALSE, message=FALSE}
cnToTake <- c("RAINDALS", "RAASIAN", "RABLACK", "RAHAWOPI", "RAWHITE", "RANOS")
patientData$Origin <- as.factor(apply(
   patientData[, cnToTake],
   1,
   function(x){
      toRet <- cnToTake[which(x=="Yes")]
      if(length(toRet)!=1){
         return("other")
      }
      if(toRet %in% c("RAINDALS", "RAHAWOPI", "RANOS")){
         return("other")
      }
      return(toRet)
   }
))
parDocToAdd <- data.frame(
    Variable="Origin",
    Description="Origin population",
    Group="Patient",
    "Sub-group"="Demographic",
    Class=class(patientData$Origin),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "Origin"
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# PD history

## Family

```{r, echo=FALSE, message=FALSE}
cnToTake <- c(
   "BIOMOMPD", "BIODADPD", "FULSIBPD",
   "HAFSIBPD", "MAGPARPD", "PAGPARPD",
   "MATAUPD", "PATAUPD", "KIDSPD"
)
patientData$PDHist <- apply(
   patientData[, cnToTake],
   1,
   function(x){
      any(x>0)
   }
)
parDocToAdd <- data.frame(
    Variable="PDHist",
    Description="PD Family History",
    Group="Patient",
    "Sub-group"="PD history",
    Class=class(patientData$PDHist),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "PDHist"
varDoc <- rbind(varDoc, parDocToAdd)
```

## Disease duration

```{r, echo=FALSE, message=FALSE}
dxdt <- merge(
    visitInfo,
    visitData[,"PDDXDT", drop=F],
    by=0
)
dxdt <- unique(dxdt[which(!is.na(dxdt$PDDXDT)),c("PATNO", "PDDXDT")])
dupPatno <- dxdt$PATNO[which(duplicated(dxdt$PATNO))]
if(length(dupPatno)>0){
    toAggr <- dxdt[which(dxdt$PATNO %in% dupPatno),]
    toAggr <- do.call(rbind, by(
        toAggr,
        toAggr$PATNO,
        function(d){
            return(data.frame(
                PATNO=d[1,"PATNO"],
                PDDXDT=min(d$PDDXDT),
                stringsAsFactors=F
            ))
        }
    ))
    dxdt <- rbind(dxdt[which(!dxdt$PATNO %in% dupPatno),], toAggr)
}
rownames(dxdt) <- dxdt$PATNO
dxdt$ENROLL_DATE <- patientData[rownames(dxdt),"ENROLL_DATE"]
dxdt$DisDuration <- as.double(dxdt$ENROLL_DATE - dxdt$PDDXDT, units="days")/365.25
patientData$DisDuration <- NA
patientData[rownames(dxdt),"DisDuration"] <- dxdt$DisDuration
parDocToAdd <- data.frame(
    Variable="DisDuration",
    Description="Duration of the disease at enrollment",
    Group="Patient",
    "Sub-group"="PD history",
    Class=class(patientData$DisDuration),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "DisDuration"
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# Unified Parkinson’s Disease Rating Scale (UPDRS)

##  Tremor dominant (TD) 

```{r, echo=FALSE, message=FALSE}
visitData$TD <- apply(
   visitData[,
      c(
         "NP2TRMR", "NP3PTRMR", "NP3PTRML",
         "NP3KTRMR", "NP3KTRML", "NP3RTARU",
         "NP3RTALU", "NP3RTARL", "NP3RTALL",
         "NP3RTALJ", "NP3RTCON"
      )
   ],
   1,
   mean
)
parDocToAdd <- data.frame(
    Variable="TD",
    Description="Tremor dominant score",
    Group="UPDRS",
    "Sub-group"="TD/PIGD",
    Class=class(visitData$TD),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "TD"
varDoc <- rbind(varDoc, parDocToAdd)
```

## Postural instability/gait difficulty (PIGD)

```{r, echo=FALSE, message=FALSE}
visitData$PIGD <- apply(
   visitData[,
     c(
        "NP2WALK", "NP2FREZ", "NP3GAIT", "NP3FRZGT",
        "NP3PSTBL"
      )
   ],
   1,
   mean
)
parDocToAdd <- data.frame(
    Variable="PIGD",
    Description="Postural instability/gait difficulty",
    Group="UPDRS",
    "Sub-group"="TD/PIGD",
    Class=class(visitData$PIGD),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "PIGD"
varDoc <- rbind(varDoc, parDocToAdd)
```

## TD/PIGD

```{r, echo=FALSE, message=FALSE}
visitData$TD.PIGD <- visitData$TD/visitData$PIGD
parDocToAdd <- data.frame(
    Variable="TD.PIGD",
    Description="TD/PIGD",
    Group="UPDRS",
    "Sub-group"="TD/PIGD",
    Class=class(visitData$TD.PIGD),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- "TD.PIGD"
varDoc <- rbind(varDoc, parDocToAdd)
```

## TD/PIGD

```{r, echo=FALSE, message=FALSE}
newVar <- "TD.PIGD.Cat"
visitData[,newVar] <- as.factor(ifelse(
   visitData$TD==0 & visitData$PIGD==0,
   "Indeterminate",
   ifelse(
      visitData$PIGD==0,
      "TD",
      ifelse(
         visitData$TD.PIGD >= 1.15,
         "TD",
         ifelse(
            visitData$TD.PIGD <= 0.9,
            "PIGD",
            "Indeterminate"
         )
      )
   )
))
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="TD/PIGD category",
    Group="UPDRS",
    "Sub-group"="TD/PIGD",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## UPDRS1

```{r, echo=FALSE, message=FALSE}
indScores1 <- c(
   "NP1COG", "NP1HALL", "NP1DPRS", "NP1ANXS",
   "NP1APAT", "NP1DDS", "NP1SLPN", "NP1SLPD",
   "NP1PAIN", "NP1URIN", "NP1CNST", "NP1LTHD",
   "NP1FATG"
)
if(any(!indScores1 %in% colnames(visitData))){
   print(setdiff(indScores1, colnames(visitData)))
   stop("not all the updrs1 scores")
}
visitData$UPDRS1 <- apply(
   visitData[,indScores1],
   1,
   sum
)
```

## UPDRS2

```{r, echo=FALSE, message=FALSE}
indScores2 <- c(
   "NP2SPCH", "NP2SALV", "NP2SWAL", "NP2EAT",
   "NP2DRES", "NP2HYGN", "NP2HWRT", "NP2HOBB",
   "NP2TURN", "NP2TRMR", "NP2RISE", "NP2WALK",
   "NP2FREZ"
)
if(any(!indScores2 %in% colnames(visitData))){
   print(setdiff(indScores2, colnames(visitData)))
   stop("not all the updrs2 scores")
}
visitData$UPDRS2 <- apply(
   visitData[,indScores2],
   1,
   sum
)
```

## UPDRS3

```{r, echo=FALSE, message=FALSE}
indScores3 <- c(
   "NP3SPCH", "NP3FACXP", "NP3RIGN", "NP3RIGRU",
   "NP3RIGLU", "PN3RIGRL", "NP3RIGLL", "NP3FTAPR",
   "NP3FTAPL", "NP3HMOVR", "NP3HMOVL",
   "NP3PRSPR", "NP3PRSPL", "NP3TTAPR",
   "NP3TTAPL", "NP3LGAGR", "NP3LGAGL",
   "NP3RISNG", "NP3GAIT", "NP3FRZGT", "NP3PSTBL",
   "NP3POSTR", "NP3BRADY", "NP3PTRMR",
   "NP3PTRML", "NP3KTRMR", "NP3KTRML",
   "NP3RTARU", "NP3RTALU", "NP3RTARL",
   "NP3RTALL", "NP3RTALJ", "NP3RTCON"
)
if(any(!indScores3 %in% colnames(visitData))){
   print(setdiff(indScores3, colnames(visitData)))
   stop("not all the updrs3 scores")
}
visitData$UPDRS3 <- apply(
   visitData[,indScores3],
   1,
   sum
)
## UPDRS
visitData$UPDRS <- apply(
   visitData[,c("UPDRS1", "UPDRS2", "UPDRS3")],
   1,
   sum
)
##
parDocToAdd <- data.frame(
    Variable=c("UPDRS1", "UPDRS2", "UPDRS3", "UPDRS"),
    Description=c(
        "UPDRS I: evaluation of mentation, behavior, and mood",
        "UPDRS II: self-evaluation of the activities of daily life (ADLs)",
        "UPDRS III: clinician-scored monitored motor evaluation",
        "Unified Parkinson’s Disease Rating Scale"
    ),
    Group=rep("UPDRS", 4),
    "Sub-group"=c("UPDRS1", "UPDRS2", "UPDRS3 pre-treatment", "UPDRS"),
    Class=rep("numeric", 4),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- c("UPDRS1", "UPDRS2", "UPDRS3", "UPDRS")
varDoc <- rbind(varDoc, parDocToAdd)
```

## UPDRS3 post treatment

```{r, echo=FALSE, message=FALSE}
indScores3 <- paste(
   indScores3,
   ".post",
   sep=""
)
if(any(!indScores3 %in% colnames(visitData))){
   print(setdiff(indScores3, colnames(visitData)))
   stop("not all the updrs3 postTr scores")
}
newVar <- "UPDRS3.post"
visitData[,newVar] <- apply(
   visitData[,indScores3],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="UPDRS3 post-treatment",
    Group="UPDRS",
    "Sub-group"="UPDRS3 post-treatment",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## UPDRS4

```{r, echo=FALSE, message=FALSE}
indScores4 <- c(
    "NP4WDYSK", "NP4DYSKI", "NP4OFF",
    "NP4FLCTI", "NP4FLCTX", "NP4DYSTN"

)
if(any(!indScores4 %in% colnames(visitData))){
   print(setdiff(indScores4, colnames(visitData)))
   stop("not all the updrs4 scores")
}
newVar <- "UPDRS4"
visitData[,newVar] <- apply(
   visitData[,indScores4],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="UPDRS IV: complications of therapy",
    Group="UPDRS",
    "Sub-group"="UPDRS4",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# Orienting features according to dominant side

```{r, echo=FALSE, message=FALSE}
sidedPar <- data.frame(
    left=c(
        "MSLARSP", "MSLLRSP", "COFNLRSP", "COHSLRSP", "SENLARSP", "SENLLRSP", 
        "RFLLARSP", "RFLLLRSP", "PLRLRSP", "NP3FTAPL", "NP3HMOVL", "NP3PRSPL", 
        "NP3TTAPL", "NP3LGAGL", "NP3PTRML", "NP3KTRML", "NP3RTALU", "NP3RTALL",
        "NP3RIGLU", "NP3RIGLL",
        "NP3FTAPL.post", "NP3HMOVL.post", "NP3PRSPL.post", 
        "NP3TTAPL.post", "NP3LGAGL.post", "NP3PTRML.post",
        "NP3KTRML.post", "NP3RTALU.post", "NP3RTALL.post",
        "NP3RIGLU.post", "NP3RIGLL.post"
    ),
    right=c(
        "MSRARSP", "MSRLRSP", "COFNRRSP", "COHSRRSP", "SENRARSP", "SENRLRSP", 
        "RFLRARSP", "RFLRLRSP", "PLRRRSP","NP3FTAPR", "NP3HMOVR", "NP3PRSPR", 
        "NP3TTAPR", "NP3LGAGR", "NP3PTRMR", "NP3KTRMR", "NP3RTARU", "NP3RTARL",
        "NP3RIGRU", "PN3RIGRL",
        "NP3FTAPR.post", "NP3HMOVR.post", "NP3PRSPR.post", 
        "NP3TTAPR.post", "NP3LGAGR.post", "NP3PTRMR.post",
        "NP3KTRMR.post", "NP3RTARU.post", "NP3RTARL.post",
        "NP3RIGRU.post", "PN3RIGRL.post"
    ),
    stringsAsFactors=F
)
sidedPar <- cbind(
    sidedPar,
    varDoc[sidedPar$left, c("Description", "Group", "Sub-group", "Class")]
)
row.names(sidedPar) <- c()
sidedPar$Description <- sub(" [lL]eft", "", sidedPar$Description)
sidedPar$Description <- sub("LUE$", "UE", sidedPar$Description)
sidedPar$Description <- sub("LLE$", "LE", sidedPar$Description)
sidedPar$Description <- sub(" -$", "", sidedPar$Description)
sidedPar$Description <- sub("[bd] ", " ", sidedPar$Description)
kable(sidedPar)
sidedPar$Source <- "Derived"
```

## Contralateral

```{r, echo=FALSE, message=FALSE}
rn <- c()
if(exists("varToAdd")) rm(varToAdd)
for(i in 1:nrow(sidedPar)){
    lvar <- sidedPar$left[i]
    rvar <- sidedPar$right[i]
    lval <- visitData[,lvar]
    rval <- visitData[,rvar]
    if(sidedPar$Class[i]=="factor"){
        lval <- as.character(lval)
        rval <- as.character(rval)
    }
    domSide <- visitData[
        paste(vPatient(visitData), "SC", sep="."),
        "DOMSIDE"
    ]
    newVar <- do.call(
        function(x, y) paste(x[which(x==y)], collapse=""),
        strsplit(c(lvar, rvar), split="")
    )
    if(newVar == "3RIGL"){
        newVar <- "NP3RIGL"
    }
    if(newVar == "3RIGL.post"){
        newVar <- "NP3RIGL.post"
    }
    newVar <- paste0(newVar, "_CL")
    rn <- c(rn, newVar)
    nval <- ifelse(
        patientData[vPatient(visitData), "ENROLL_CAT"]=="HC",
        NA,
        ifelse(
            domSide=="Right", #2
            lval,
            ifelse(
                domSide=="Left", #1
                rval,
                ifelse(
                    domSide=="Symmetric", #3
                    NA,
                    NA
                )
            )
        )
    )
    if(exists("varToAdd")){
        varToAdd <- data.frame(varToAdd, nval, stringsAsFactors=T)
    }else{
        varToAdd <- nval
    }
}
colnames(varToAdd) <- rn
parDocToAdd <- data.frame(
    "Variable"=rn,
    sidedPar[,c("Description", "Group", "Sub-group", "Class", "Source")],
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- rn
parDocToAdd$Description <- paste(parDocToAdd$Description, "- Contralateral")
visitData <- cbind(visitData, varToAdd)
varDoc <- rbind(varDoc, parDocToAdd)
##
indScores <- c(
    "NP3FTAP_CL", "NP3HMOV_CL",
    "NP3PRSP_CL", "NP3TTAP_CL", 
    "NP3LGAG_CL", "NP3PTRM_CL", 
    "NP3KTRM_CL", "NP3RTAU_CL",
    "NP3RIGU_CL", "NP3RIGL_CL",
    "NP3RTAL_CL"
)
if(any(!indScores %in% colnames(visitData))){
   print(setdiff(indScores, colnames(visitData)))
   stop("not all the CL updrs3 scores")
}
newVar <- "SUPDRS3_CL"
visitData[,newVar] <- apply(
   visitData[,indScores],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Sub-UPDRS3 - Contralateral",
    Group="UPDRS",
    "Sub-group"="UPDRS3 pre-treatment",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
indScores <- c(
    "NP3FTAP.post_CL", "NP3HMOV.post_CL",
    "NP3PRSP.post_CL", "NP3TTAP.post_CL", 
    "NP3LGAG.post_CL", "NP3PTRM.post_CL", 
    "NP3KTRM.post_CL", "NP3RTAU.post_CL",
    "NP3RIGU.post_CL", "NP3RIGL.post_CL",
    "NP3RTAL.post_CL"
)
if(any(!indScores %in% colnames(visitData))){
   print(setdiff(indScores, colnames(visitData)))
   stop("not all the CL updrs3 scores")
}
newVar <- "SUPDRS3.post_CL"
visitData[,newVar] <- apply(
   visitData[,indScores],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Sub-UPDRS3 post-treatment - Contralateral",
    Group="UPDRS",
    "Sub-group"="UPDRS3 post-treatment",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Ipsilateral

```{r, echo=FALSE, message=FALSE}
rn <- c()
rm(varToAdd)
for(i in 1:nrow(sidedPar)){
    lvar <- sidedPar$left[i]
    rvar <- sidedPar$right[i]
    lval <- visitData[,lvar]
    rval <- visitData[,rvar]
    if(sidedPar$Class[i]=="factor"){
        lval <- as.character(lval)
        rval <- as.character(rval)
    }
    domSide <- visitData[
        paste(vPatient(visitData), "SC", sep="."),
        "DOMSIDE"
    ]
    newVar <- do.call(
        function(x, y) paste(x[which(x==y)], collapse=""),
        strsplit(c(lvar, rvar), split="")
    )
    if(newVar == "3RIGL"){
        newVar <- "NP3RIGL"
    }
    if(newVar == "3RIGL.post"){
        newVar <- "NP3RIGL.post"
    }
    newVar <- paste0(newVar, "_IL")
    rn <- c(rn, newVar)
    nval <- ifelse(
        patientData[vPatient(visitData), "ENROLL_CAT"]=="HC",
        NA,
        ifelse(
            domSide=="Right", #2
            rval,
            ifelse(
                domSide=="Left", #1
                lval,
                ifelse(
                    domSide=="Symmetric", #3
                    NA,
                    NA
                )
            )
        )
    )
    if(exists("varToAdd")){
        varToAdd <- data.frame(varToAdd, nval, stringsAsFactors=T)
    }else{
        varToAdd <- nval
    }
}
colnames(varToAdd) <- rn
parDocToAdd <- data.frame(
    "Variable"=rn,
    sidedPar[,c("Description", "Group", "Sub-group", "Class", "Source")],
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- rn
parDocToAdd$Description <- paste(parDocToAdd$Description, "- Ipsilateral")
visitData <- cbind(visitData, varToAdd)
varDoc <- rbind(varDoc, parDocToAdd)
##
indScores <- c(
    "NP3FTAP_IL", "NP3HMOV_IL",
    "NP3PRSP_IL", "NP3TTAP_IL", 
    "NP3LGAG_IL", "NP3PTRM_IL", 
    "NP3KTRM_IL", "NP3RTAU_IL",
    "NP3RIGU_IL", "NP3RIGL_IL",
    "NP3RTAL_IL"
)
if(any(!indScores %in% colnames(visitData))){
   print(setdiff(indScores, colnames(visitData)))
   stop("not all the CL updrs3 scores")
}
newVar <- "SUPDRS3_IL"
visitData[,newVar] <- apply(
   visitData[,indScores],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Sub-UPDRS3 - Ipsilateral",
    Group="UPDRS",
    "Sub-group"="UPDRS3 pre-treatment",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
indScores <- c(
    "NP3FTAP.post_IL", "NP3HMOV.post_IL",
    "NP3PRSP.post_IL", "NP3TTAP.post_IL", 
    "NP3LGAG.post_IL", "NP3PTRM.post_IL", 
    "NP3KTRM.post_IL", "NP3RTAU.post_IL",
    "NP3RIGU.post_IL", "NP3RIGL.post_IL",
    "NP3RTAL.post_IL"
)
if(any(!indScores %in% colnames(visitData))){
   print(setdiff(indScores, colnames(visitData)))
   stop("not all the CL updrs3 scores")
}
newVar <- "SUPDRS3.post_IL"
visitData[,newVar] <- apply(
   visitData[,indScores],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Sub-UPDRS3 post-treatment - Ipsilateral",
    Group="UPDRS",
    "Sub-group"="UPDRS3 post-treatment",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# Non-Motor assessments

## Benton judgment of line oriention test (BJLOT)

```{r, echo=FALSE, message=FALSE}
newVar <- "BJLOT"
# Only on 15 parameters
visitData[,newVar] <- apply(
   visitData[,paste("BJLOT", 1:30, sep="")],
   1,
   function(x){
      toRet <- sum(x[seq(1, 29, by=2)]=="Correct")
      if(is.na(toRet)){
         toRet <- sum(x[seq(2, 30, by=2)]=="Correct")
      }
      return(toRet)
   }
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Benton judgment of line oriention test (BJLOT)",
    Group="Non-motor",
    "Sub-group"="BJLOT",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Epworth sleepiness scale (ESS)

```{r, echo=FALSE, message=FALSE}
newVar <- "ESS"
visitData[,newVar] <- apply(
   visitData[,paste("ESS", 1:8, sep="")],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Epworth sleepiness scale (ESS)",
    Group="Non-motor",
    "Sub-group"="ESS",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "Sleepy"
visitData[,newVar] <- visitData$ESS >= 10
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Sleepy: ESS >= 10",
    Group="Non-motor",
    "Sub-group"="ESS",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Geriatric depression scale (GDS)

```{r, echo=FALSE, message=FALSE}
newVar <- "GDS"
visitData[,newVar] <- apply(
   visitData[,c(
      "GDSSATIS", "GDSGSPIR", "GDSHAPPY",
      "GDSALIVE", "GDSENRGY"
   )],
   1,
   function(x) sum(x==0)
) + apply(
   visitData[,c(
      "GDSDROPD", "GDSEMPTY", "GDSBORED",
      "GDSAFRAD", "GDSHLPLS", "GDSHOME",
      "GDSMEMRY", "GDSWRTLS", "GDSHOPLS",
      "GDSBETER"
   )],
   1,
   function(x) sum(x==1)
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Geriatric depression scale (GDS)",
    Group="Non-motor",
    "Sub-group"="GDS",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "Depressed"
visitData[,newVar] <- visitData$GDS >= 5
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Depressed: GDS >= 5",
    Group="Non-motor",
    "Sub-group"="GDS",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Hopkins verbal learning test (HVLT)

```{r, echo=FALSE, message=FALSE}
newVar <- "HVLTIR"
visitData[,newVar] <- apply(
   visitData[, paste("HVLTRT", 1:3, sep="")],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Hopkins verbal learning test (HVLT) - Immediate Recall",
    Group="Non-motor",
    "Sub-group"="HVLT",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)

## HVLT Discrimination Recognition
newVar <- "HVLTDR"
visitData[,newVar] <- visitData$HVLTREC - visitData$HVLTFPRL + visitData$HVLTFPUN
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Hopkins verbal learning test - Discrimination Recognition",
    Group="Non-motor",
    "Sub-group"="HVLT",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Letter and number sequencing (LNS)

```{r, echo=FALSE, message=FALSE}
newVar <- "LNS"
visitData[,newVar] <- apply(
   visitData[, paste("LNS", rep(1:7, each=3), c("A", "B", "C"), sep="")],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Letter and number sequencing (LNS)",
    Group="Non-motor",
    "Sub-group"="LNS",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Montreal cognitive assessment (MOCA)

```{r, echo=FALSE, message=FALSE}
newVar <- "MOCA"
visitData[,newVar] <- apply(
   visitData[,c(
      "MCAALTTM",
      "MCACUBE", "MCACLCKC", "MCACLCKN",
      "MCACLCKH", "MCALION", "MCARHINO",
      "MCACAMEL", "MCAFDS", "MCABDS",
      "MCAVIGIL", "MCASER7", "MCASNTNC", "MCAVF",
      "MCAABSTR", "MCAREC1", "MCAREC2",
      "MCAREC3", "MCAREC4", "MCAREC5",
      "MCADATE", "MCAMONTH", "MCAYR",
      "MCADAY", "MCAPLACE", "MCACITY"
   )],
   1,
   sum
)
visitData[,newVar] <- ifelse(
   patientData[vPatient(visitData), "EDUCYRS"] <= 12 & visitData[,newVar] < 30,
   visitData[,newVar] + 1,
   visitData[,newVar]
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Montreal cognitive assessment (MOCA)",
    Group="Non-motor",
    "Sub-group"="MOCA",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
   )
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Questionnaire for Impulsive-Compulsive Disorders in PD (QUIP)

```{r, echo=FALSE, message=FALSE}
newVar <- "QUIP"
visitData[,newVar] <- apply(
   visitData[,c(
      "CNTRLGMB", "TMGAMBLE"
   )],
   1,
   function(x) any(x=="Yes")
) + apply(
   visitData[,c(
      "CNTRLSEX", "TMSEX"
   )],
   1,
   function(x) any(x=="Yes")
) + apply(
   visitData[,c(
      "CNTRLBUY", "TMBUY"
   )],
   1,
   function(x) any(x=="Yes")
) + apply(
   visitData[,c(
      "CNTRLEAT", "TMEAT"
   )],
   1,
   function(x) any(x=="Yes")
) + apply(
   visitData[,c(
      "TMTORACT", "TMTMTACT", "TMTRWD"
   )],
   1,
   function(x) sum(x=="Yes")
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Questionnaire for Impulsive-Compulsive Disorders in PD (QUIP)",
    Group="Non-motor",
    "Sub-group"="QUIP",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## REM (Rapid eye movement) sleep behavior disorder (RBD)

```{r, echo=FALSE, message=FALSE}
newVar <- "RBD"
visitData[,newVar] <- apply(
   visitData[,c(
      "DRMVIVID", "DRMAGRAC", "DRMNOCTB",
      "SLPLMBMV", "SLPINJUR", "DRMVERBL",
      "DRMFIGHT", "DRMUMV", "DRMOBJFL",
      "MVAWAKEN", "DRMREMEM", "SLPDSTRB"
   )],
   1,
   function(x) sum(x=="Yes")
) + apply(
   visitData[,c(
      "STROKE", "HETRA", "PARKISM", "RLS",
      "NARCLPSY", "DEPRS", "EPILEPSY", "BRNINFM",
      "CNSOTH"
   )],
   1,
   function(x) any(x=="Yes")
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="REM Sleep Behavior disorder (RBD)",
    Group="Non-motor",
    "Sub-group"="RBD",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "RBD.pos"
visitData[,newVar] <- visitData$RBD >=5
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="RBD Positive: RBD >= 5",
    Group="Non-motor",
    "Sub-group"="RBD",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
   )
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Scales for outcomes in Parkinson’s disease-autonomic (SCOPA-AUT)

```{r, echo=FALSE, message=FALSE}
newVar <- "SCOPA"
visitData[,newVar] <- apply(
   visitData[, paste("SCAU", 1:21, sep="")],
   1,
   function(x){
      x <- ifelse(
         x=="never", 0,
         ifelse(
            x=="sometimes", 1,
            ifelse(
               x=="regularly", 2,
               ifelse(
                  x=="often", 3,
                  ifelse(
                     x=="use catheter",
                     9,
                     NA
                  )
               )
            )
         )
      )
      sum(ifelse(x==9, 3, x), na.rm=T)
   }
) + apply(
   visitData[, paste("SCAU", 22:25, sep="")],
   1,
   function(x){
      x <- ifelse(
         x=="never", 0,
         ifelse(
            x=="sometimes", 1,
            ifelse(
               x=="regularly", 2,
               ifelse(
                  x=="often", 3,
                  ifelse(
                     x=="not applicable",
                     9,
                     NA
                  )
               )
            )
         )
      )
      sum(ifelse(x==9, 0, x), na.rm=T)
   }
)
#### !!! ####
## Only 2 patients with values for all SCAU22 --> SCAU25
## ===> do not take NA into account
#### !!! ####
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Scales for outcomes in Parkinson’s disease-autonomic (SCOPA-AUT)",
    Group="Non-motor",
    "Sub-group"="SCOPA-AUT",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Semantic Fluency

```{r, echo=FALSE, message=FALSE}
newVar <- "SFT"
visitData[,newVar] <- apply(
   visitData[, c("VLTANIM", "VLTVEG", "VLTFRUIT")],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Semantic fluency",
    Group="Non-motor",
    "Sub-group"="Semantic fluency",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## State Trait Anxiety Total Score (STA)

```{r, echo=FALSE, message=FALSE}
newVar <- "STA"
withOriScore <- c(
   3, 4, 6, 7, 9, 12, 13, 14, 17, 18, 22, 24, 25, 28, 29,
   31, 32, 35, 37, 38, 40
)
visitData[,newVar] <- apply(
   visitData[, paste("STAIAD", withOriScore, sep="")],
   1,
   sum
) + apply(
   visitData[, paste("STAIAD", setdiff(1:40, withOriScore), sep="")],
   1,
   function(x){
      x <- (4:1)[x]
      sum(x)
   }
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="State Trait Anxiety Total Score",
    Group="Non-motor",
    "Sub-group"="STA",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)

## STAI - State Subscore
newVar <- "STAI.State"
withOriScore <- c(
   3, 4, 6, 7, 9, 12, 13, 14, 17, 18
)
visitData[,newVar] <- apply(
   visitData[, paste("STAIAD", withOriScore, sep="")],
   1,
   sum
) + apply(
   visitData[, paste("STAIAD", setdiff(1:20, withOriScore), sep="")],
   1,
   function(x){
      x <- (4:1)[x]
      sum(x)
   }
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="STAI - State Subscore",
    Group="Non-motor",
    "Sub-group"="STA",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)

## STAI - Trait Subscore
newVar <- "STAI.Trait"
withOriScore <- c(
   22, 24, 25, 28, 29,
   31, 32, 35, 37, 38, 40
)
visitData[,newVar] <- apply(
   visitData[, paste("STAIAD", withOriScore, sep="")],
   1,
   sum
) + apply(
   visitData[, paste("STAIAD", setdiff(21:40, withOriScore), sep="")],
   1,
   function(x){
      x <- (4:1)[x]
      sum(x)
   }
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="STAI - Trait Subscore",
    Group="Non-motor",
    "Sub-group"="STA",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## University of Pennsylvania Smell ID Test (UPSIT)

```{r, echo=FALSE, message=FALSE}
newVar <- "UPSIT"
visitData[,newVar] <- apply(
   visitData[, paste("UPSITBK", 1:4, sep="")],
   1,
   sum
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="University of Pennsylvania Smell ID Test (UPSIT)",
    Group="Non-motor",
    "Sub-group"="UPSIT",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Mild Cognitive Impairment

```{r, echo=FALSE, message=FALSE}
newVar <- "MCI"
cogScore <- data.frame(
    HVLTIR=visitData$HVLTIR <= 35,
    HVLTDR=visitData$HVLTDR <= 35,
    BJLOT=visitData$BJLOT <= 6,
    LNS=visitData$LNS <= 6,
    SFT=visitData$SFT <= 35,
    SDM=visitData$DVT_SDM <= 35
)
cogScore <- apply(
    cogScore,
    1,
    function(x) sum(x)
)
visitData[,newVar] <- (
    visitData$COGDECLN=="Yes" &
    cogScore >= 2 &
    visitData$FNCDTCOG=="No"
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Mild Cognitive Impairment",
    Group="Non-motor",
    "Sub-group"="Cognitive categorization",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# Concomitant medications

```{r}
colnames(cmed) <- sub("^EVENT_ID$", "RECORD_EVENT", colnames(cmed))
cmed <- merge(cmed, visitInfo, by="PATNO", all=FALSE)
cmed <- cmed[which(!is.na(cmed$INFODT)),]
cmed <- cmed[which(
   (is.na(cmed$STARTDT) | cmed$STARTDT < cmed$INFODT) &
      (is.na(cmed$STOPDT) | cmed$STOPDT>= cmed$INFODT),
),]
agcmed <- do.call(rbind, by(
   cmed,
   paste(cmed$PATNO, cmed$EVENT_ID),
   function(x){
      return(data.frame(
         x[1, c("PATNO", "EVENT_ID")],
         # LEDD=ifelse(all(is.na(x$LEDD_num)), NA, sum(x$LEDD_num, na.rm=TRUE)),
         "MED_WITH_LEDD"=sum(!is.na(x$LEDD_num)),
         "NON_DIS_MED"=sum(x$DISMED==0),
         "DIS_MED"=sum(x$DISMED!=0),
         "NON_PD_MOTOR_MED"=sum(x$PD_MOTOR_MED==0),
         "PD_MOTOR_MED"=sum(x$PD_MOTOR_MED!=0)
         # "WITH_LDX"=any(x$LEDD_str=="LDx")
      ))
   }
))
rownames(agcmed) <- paste(agcmed$PATNO, agcmed$EVENT_ID, sep=".")
agcmed <- agcmed[rownames(visitData),]
rownames(agcmed) <- rownames(visitData)
visitData <- cbind(
  visitData,
  agcmed[,setdiff(colnames(agcmed), c("PATNO", "EVENT_ID"))]
)
parDocToAdd <- data.frame(
    Variable=c(
      # "LEDD",
      "MED_WITH_LEDD",
      "NON_DIS_MED",
      "DIS_MED",
      "NON_PD_MOTOR_MED",
      "PD_MOTOR_MED"
    ),
    Description=c(
      # "Sum of L-dopa equivalent daily dose (LEDD)",
      "Number of medications with a LEDD value",
      "Number of non PD medications",
      "Number of PD medications",
      "Number of medications not affecting PD Motor assessments",
      "Number of medications affecting PD Motor assessments"
    ),
    Group="Medical history",
    "Sub-group"="PD medication",
    Class="numeric",
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
   )
rownames(parDocToAdd) <- parDocToAdd$Variable
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# Biological data

## t-tau/Abeta 1-42

```{r, echo=FALSE, message=FALSE}
newVar <- "tTau.Abeta"
visitData[,newVar] <- visitData$"Total tau"/visitData$"Abeta 42"
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="t-tau/Abeta 1-42",
    Group="Biological",
    "Sub-group"="Cerebrospinal fluid",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## p-tau/Abeta 1-42

```{r, echo=FALSE, message=FALSE}
newVar <- "pTau.Abeta"
visitData[,newVar] <- visitData$"p-Tau181P"/visitData$"Abeta 42"
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="p-tau/Abeta 1-42",
    Group="Biological",
    "Sub-group"="Cerebrospinal fluid",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## p-tau/t-tau

```{r, echo=FALSE, message=FALSE}
newVar <- "pTau.tTau"
visitData[,newVar] <- visitData$"p-Tau181P"/visitData$"Total tau"
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="p-tau/t-tau",
    Group="Biological",
    "Sub-group"="Cerebrospinal fluid",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

<!-------------------------------------------------->
<!-------------------------------------------------->

# DatScan

## Contralateral

```{r, echo=FALSE, message=FALSE}
newVar <- "CAUDATE_CL"
visitData[,newVar] <- ifelse(
   patientData[vPatient(visitData), "ENROLL_CAT"]=="HC",
   apply(visitData[, paste("CAUDATE", c("R", "L"), sep="_")], 1, mean),
   ifelse(
      visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Right", #2
      visitData$CAUDATE_L,
      ifelse(
         visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Left", #1
         visitData$CAUDATE_R,
         ifelse(
            visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Symmetric", #3
            apply(visitData[, paste("CAUDATE", c("R", "L"), sep="_")], 1, mean),
            NA
         )
      )
   )
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="CAUDATE Contralateral",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "PUTAMEN_CL"
visitData[,newVar] <- ifelse(
   patientData[vPatient(visitData), "ENROLL_CAT"]=="HC",
   apply(visitData[, paste("PUTAMEN", c("R", "L"), sep="_")], 1, mean),
   ifelse(
      visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Right", #2
      visitData$PUTAMEN_L,
      ifelse(
         visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Left", #1
         visitData$PUTAMEN_R,
         ifelse(
            visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Symmetric", #3
            apply(visitData[, paste("PUTAMEN", c("R", "L"), sep="_")], 1, mean),
            NA
            )
         )
      )
   )
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="PUTAMEN Contralateral",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Ipsilateral

```{r, echo=FALSE, message=FALSE}
newVar <- "CAUDATE_IL"
visitData[,newVar] <- ifelse(
   patientData[vPatient(visitData), "ENROLL_CAT"]=="HC",
   apply(visitData[, paste("CAUDATE", c("R", "L"), sep="_")], 1, mean),
   ifelse(
      visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Left", #2
      visitData$CAUDATE_L,
      ifelse(
         visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Right", #1
         visitData$CAUDATE_R,
         ifelse(
            visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Symmetric", #3
            apply(visitData[, paste("CAUDATE", c("R", "L"), sep="_")], 1, mean),
            NA
            )
         )
      )
   )
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="CAUDATE Ipsilateral",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "PUTAMEN_IL"
visitData[,newVar] <- ifelse(
   patientData[vPatient(visitData), "ENROLL_CAT"]=="HC",
   apply(visitData[, paste("PUTAMEN", c("R", "L"), sep="_")], 1, mean),
   ifelse(
      visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Left", #2
      visitData$PUTAMEN_L,
      ifelse(
         visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Right", #1
         visitData$PUTAMEN_R,
         ifelse(
            visitData[paste(vPatient(visitData), "SC", sep="."),"DOMSIDE"]=="Symmetric", #3
            apply(visitData[, paste("PUTAMEN", c("R", "L"), sep="_")], 1, mean),
            NA
            )
         )
      )
   )
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="PUTAMEN Ipsilateral",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Mean Caudate

```{r, echo=FALSE, message=FALSE}
newVar <- "CAUDATE"
visitData[,newVar] <- apply(
   visitData[,paste(newVar, c("R", "L"), sep="_")],
   1,
   mean
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Mean Caudate",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Mean Putamen

```{r, echo=FALSE, message=FALSE}
newVar <- "PUTAMEN"
visitData[,newVar] <- apply(
   visitData[,paste(newVar, c("R", "L"), sep="_")],
   1,
   mean
   )
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Mean Putamen",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Mean Striatum

```{r, echo=FALSE, message=FALSE}
newVar <- "STRIATUM"
visitData[,newVar] <- apply(
   visitData[,paste(rep(c("CAUDATE", "PUTAMEN"), each=2), rep(c("R", "L"), 2), sep="_")],
   1,
   mean
   )
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Mean Striatum",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Count Density Ratio

```{r, echo=FALSE, message=FALSE}
newVar <- "CDR"
visitData[,newVar] <- visitData$CAUDATE/visitData$PUTAMEN
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Count Density Ratio: Caudate/Putamen",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "CDR_CL"
visitData[,newVar] <- visitData$CAUDATE_CL/visitData$PUTAMEN_CL
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Count Density Ratio (CL): Caudate_CL/Putamen_CL",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "CDR_IL"
visitData[,newVar] <- visitData$CAUDATE_IL/visitData$PUTAMEN_IL
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Count Density Ratio (IL): Caudate_IL/Putamen_IL",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Asymetry Index

```{r, echo=FALSE, message=FALSE}
newVar <- "CAUDATE_AI"
visitData[,newVar] <- apply(
   visitData[,paste("CAUDATE", c("R", "L"), sep="_")],
   1,
   function(x){
      abs(x[2]-x[1])*100/mean(x)
   }
)
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Caudate Asymetry Index",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
##
newVar <- "PUTAMEN_AI"
visitData[,newVar] <- apply(
   visitData[,paste("PUTAMEN", c("R", "L"), sep="_")],
   1,
   function(x){
      abs(x[2]-x[1])*100/mean(x)
   }
   )
parDocToAdd <- data.frame(
    Variable=newVar,
    Description="Putamen Asymetry Index",
    Group="Imaging",
    "Sub-group"="DaTScan",
    Class=class(visitData[, newVar]),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- newVar
varDoc <- rbind(varDoc, parDocToAdd)
```

## Ratio (\%) to age expected value in healthy control

### DaTScan SBR in healthy controls at the screening visit

```{r, echo=FALSE, message=FALSE}
datScanFeat <- c(
   "CAUDATE_L", "CAUDATE_R", "CAUDATE_CL", "CAUDATE_IL", "CAUDATE",
   "PUTAMEN_L", "PUTAMEN_R", "PUTAMEN_CL", "PUTAMEN_IL", "PUTAMEN",
   "STRIATUM"
)
hcPat <- rownames(patientData)[which(patientData$ENROLL_CAT=="HC")]
hcScImgTab <- visitData[
    which(vPatient(visitData) %in% hcPat & vEvent(visitData)=="SC"),
    datScanFeat
]
scDate <- visitInfo[rownames(hcScImgTab), "INFODT"]
biDate <- patientData[vPatient(hcScImgTab), "BIRTHDT"]
age <- as.double(scDate - biDate, units="days")/365.25
gender <- patientData[vPatient(hcScImgTab), "SimpleGender"]
rownames(hcScImgTab) <- vPatient(hcScImgTab)

### Modeling
models <- list()
for(i in colnames(hcScImgTab)){
    x <- age
    y <- hcScImgTab[,i]
    plot(
        x, y,
        xlab="Age", ylab="SBR",
        bg=ifelse(gender=="Male", "blue", "pink"),
        pch=21,
        main=i
    )
    ##
    x.m <- x[which(gender=="Male")]
    y.m <- y[which(gender=="Male")]
    xyCor.m <- cor.test(x.m,y.m)
    rsq.m <- xyCor.m$estimate^2
    pval.m <- xyCor.m$p.value
    model.m <- lm(y.m~x.m)
    curve(
      model.m$coefficients["x.m"]*x + model.m$coefficients["(Intercept)"],
      col="blue", lwd=2,
      add=T
    )
    legend(
        "bottomleft",
        legend=as.expression(c(
            "Male",
            paste("Slope =", signif(model.m$coefficients["x.m"],2)),
            substitute(paste(R^2, "=", r), list(r=signif(rsq.m,2))),
            substitute(paste(p.value, "=", p), list(p=signif(pval.m,2))),
            paste("n =", sum(!is.na(x.m) & !is.na(y.m)))
        )),
        pch=c(21, rep(19, 4)),
        pt.bg=c("blue", rep("transparent", 4)),
        col=c("black", rep("transparent", 4)),
        cex=0.7
        #bty="n"
    )
    ##
    x.f <- x[which(gender=="Female")]
    y.f <- y[which(gender=="Female")]
    xyCor.f <- cor.test(x.f,y.f)
    rsq.f <- xyCor.f$estimate^2
    pval.f <- xyCor.f$p.value
    model.f <- lm(y.f~x.f)
    curve(
        model.f$coefficients["x.f"]*x + model.f$coefficients["(Intercept)"],
        col="pink", lwd=2,
        add=T
    )
    legend(
        "topright",
        legend=as.expression(c(
            "Female",
            paste("Slope =", signif(model.f$coefficients["x.f"],2)),
            substitute(paste(R^2, "=", r), list(r=signif(rsq.f,2))),
            substitute(paste(p.value, "=", p), list(p=signif(pval.f,2))),
            paste("n =", sum(!is.na(x.f) & !is.na(y.f)))
        )),
        pch=c(21, rep(19, 4)),
        pt.bg=c("pink", rep("transparent", 4)),
        col=c("black", rep("transparent", 4)),
        cex=0.7
        #bty="n"
    )
    ##
    models <- c(models,list(list(m=model.m, f=model.f)))
}
names(models) <- colnames(hcScImgTab)
```

### Computing ratio (\%) to age expected value for each patient and each visit

```{r, echo=FALSE, message=FALSE}
toConv <- visitData[, datScanFeat]
scDate <- visitInfo[rownames(visitData), "INFODT"]
biDate <- patientData[vPatient(visitData), "BIRTHDT"]
age <- as.double(scDate - biDate, units="days")/365.25
age <- ifelse(
   is.na(age),
   patientData[vPatient(visitData), "ENROLL_AGE"]+ifelse(
      vEvent(visitData) == "SC", 0,
      ifelse(
         vEvent(visitData)=="V04", 1,
         ifelse(
            vEvent(visitData)=="V06", 2,
            ifelse(
               vEvent(visitData)=="V10", 4,
               NA
            )
         )
      )
   ),
   age
)
gender <- patientData[vPatient(visitData), "SimpleGender"]
percAgeExp <- c()
for(i in colnames(toConv)){
   expVal <- ifelse(
      gender=="Male",
      models[[i]]$m$coefficients["x.m"]*age + models[[i]]$m$coefficients["(Intercept)"],
      models[[i]]$f$coefficients["x.f"]*age + models[[i]]$f$coefficients["(Intercept)"]
   )
   percAgeExp <- cbind(percAgeExp, toConv[,i]*100/expVal)
}
colnames(percAgeExp) <- paste(colnames(toConv), " (%)", sep="")
rownames(percAgeExp) <- rownames(toConv)
visitData <- cbind(visitData, percAgeExp[rownames(visitData),])
parDocToAdd <- data.frame(
    Variable=colnames(percAgeExp),
    Description=paste(colnames(toConv), "- ratio to age expected value in HC"),
    Group=rep("Imaging", ncol(toConv)),
    "Sub-group"=rep("DaTScan", ncol(toConv)),
    Class=rep("numeric", ncol(toConv)),
    Source="Derived",
    stringsAsFactors=F,
    check.names=F
)
rownames(parDocToAdd) <- colnames(percAgeExp)
varDoc <- rbind(varDoc, parDocToAdd)
```

# Saving the data

```{r, echo=FALSE, message=FALSE}
varDoc <- varDoc[,setdiff(colnames(varDoc), "Variable")]
toSave <- c(
    "dumpDate",
    "visitDoc", "visitInfo", "scheduledVisits",
    "varDoc", "patientData", "visitData"
)
f <- "PPMI-Derived.rda"
save(list=toSave, file=f)
```

The following objects were saved in the *`r mdSan(f)`* file:

```{r, echo=FALSE, message=FALSE, results="asis"}
cat(paste("*", toSave), sep="\n")
```
