---
title: "PPMI: data usage"
author: "Patrice Godard"
date: "`r format(Sys.time(), '%B %d %Y')`"
package: "PPMI (version `r packageVersion('PPMI')`)"
abstract: "This document shows how to use the PPMI R package to use data from the Parkinson's Progression Markers Initiative (PPMI)."
vignette: >
    %\VignetteIndexEntry{PPMI data usage}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
output:
    BiocStyle::html_document:
        toc: true
---

```{r}
library(PPMI)
```


```{r, echo=FALSE, message=FALSE}
null <- gc()
try(null <- dev.off(), silent=T)
rm(null)
library(knitr)
library(RColorBrewer)
dumpDate <- as.Date(packageDescription("PPMI")$Date)
#######
beeswarm <- function(
    formula,
    data,
    corral="density",
    corralWidth=0.8,
    do.plot=TRUE,
    ...
){
    if(corral!="density"){
        toRet <- do.call(beeswarm::beeswarm, c(
            list(
                formula=formula,
                data=data,
                do.plot=F,
                corral=corral,
                corralWidth=corralWidth
            ),
            list(...)
        ))
    }else{
        toRet <- do.call(beeswarm::beeswarm, c(
            list(
                formula=formula,
                data=data,
                do.plot=F
            ),
            list(...)
        ))
        toRet <- do.call(rbind, by(
            toRet,
            toRet$x.orig,
            function(d){
                cpos <- round(mean(d$x, na.rm=T))
                xos <- d$x - cpos
                div <- max(abs(xos), na.rm=T)/(corralWidth/2)
                d$x <- cpos + (xos/div)
                return(d)
            }
        ))
    }
    if(do.plot){
        points(
            toRet$x, toRet$y,
            pch=toRet$pch,
            col=toRet$col, bg=toRet$bg
        )
    }
    invisible(toRet)
}
#######
rebox <- function(bp, boxwex=0.8, ...){
    sp <- boxwex/2
    for(i in 1:ncol(bp$stats)){
        y <- bp$stats[2:4, i]
        segments(
            x0=rep(i-sp, length(y)),
            x1=rep(i+sp, length(y)),
            y0=y,
            y1=y,
            ...
        )
    }
}
```

# Introduction

## Background

In the field of Parkinson’s disease (PD) therapeutics, the ultimate goal is to
develop disease-modifying treatments that slow, prevent or reverse the
underlying disease process. Validated biomarkers of disease progression would
dramatically accelerate PD therapeutics research.
Current progression biomarkers, however, are not optimal and are not
fully validated.
(source: [PPMI website](http://www.ppmi-info.org/about-ppmi/))

## Rationale for PPMI

PPMI (Parkinson's Progression Markers Initiative)
is an observational clinical study to verify progression markers in
Parkinson’s disease. PPMI has emerged as a model for following multiple
cohorts of significant interest and is being conducted at a network of
clinical sites around the world. The study is designed to establish a
comprehensive set of clinical, imaging and biosample data that will be used
to define biomarkers of PD progression.
Once these biomarkers are defined, they can be used in therapeutic studies,
which is the ultimate goal.
(source: [PPMI website](http://www.ppmi-info.org/about-ppmi/))

PPMI will follow standardized data acquisition protocols to ensure that
tests and assessments conducted at multiple sites and across multiple cohorts
can be pooled in centralized databases and repositories.
The clinical, imaging and biologic data will be easily accessible
to researchers in real time through
the [PPMI website](http://www.ppmi-info.org/about-ppmi/).
The biological samples collected throughout the course of PPMI will
be stored in a central repository that will be accessible to any scientist
with promising biomarker leads for the purposes of verifying initial
results and assessing correlations to clinical outcomes and other biomarkers.
(source: [PPMI website](http://www.ppmi-info.org/about-ppmi/))

## Data source

Data were dumped on
`r format(dumpDate, "%A %B %d %Y")` from the
[PPMI repository](https://www.ppmi-info.org/access-data-specimens/download-data/). They
were then preprocessed and integrated and derived variables were computed as
recommended by the PPMI.

This document shows how to use the PPMI R package to use data
from the Parkinson's Progression Markers Initiative (PPMI).

# Patient cohorts

The PPMI is focused on
`r length(setdiff(unique(patientData$ENROLL_CAT), NA))` patient cohorts
which can be retrieved using the `listCohorts` function:

```{r, echo=FALSE, message=FALSE}
toplot <- listCohorts()
kable(toplot, row.names=F)
```

These cohorts are described in this
[PPMI webpage](http://www.ppmi-info.org/study-design/study-cohorts/).

# Variable documentation

## Groups of variables

Variables have been classified in groups and sub-groups. In total, there are
`r nrow(varDoc)` variables belonging to
`r nrow(unique(varDoc[,c("Group", "Sub-group")]))` sub-groups in
`r length(unique(varDoc$Group))` groups.
This information is summarized by the `variableGroupSummary` function.

```{r, echo=FALSE, message=FALSE}
kable(variableGroupSummary(), row.names=F)
```

## Classes and source of variables

Variables can be of different classes as shown in the following table.
*Character* variables correspond to comments precising patient status or
treatment.

```{r, echo=FALSE, message=FALSE}
toplot <- as.data.frame(table(varDoc$Class))
colnames(toplot) <- c("Class", "Number of variables")
kable(toplot, row.names=F)
```

Variables have been either directly extracted from files provided by
the PPMI or derived according to PPMI recommendations:

```{r, echo=FALSE, message=FALSE}
toplot <- as.data.frame(table(varDoc$Source))
colnames(toplot) <- c("Source", "Number of variables")
kable(toplot, row.names=F)
```

## Finding variables

Documentation of variables is available in the `varDoc` object.

```{r}
head(varDoc)
```

Variables corresponding to groups or sub-groups can be found with the
`findVarByGroup` function.

```{r}
findVarByGroup("Plasma")
```

Variables can also be search directly according to their names or description:

```{r}
findVariable("enrollment")
```

# Patient variables

Among all availables variables, `r ncol(patientData)` were assessed only once,
during the recruitment. They belong to the group of variable named *Patient*.
These data are available in the `patientData` object.

```{r}
dim(patientData)
patientData[1:5,1:5]
```

# Follow-up variables

## Visits

Patients condition is assessed during different visits. The possible visits are
described in the `visitDoc` table.

```{r, echo=FALSE, message=FALSE}
kable(visitDoc, row.names=F)
```

Among these visits, some are scheduled from the begining of the study.
The list of these visits is available in the `scheduledVisits` table.
The table below shows the number of patients which were followed-up at each
visit when the data were downloaded (`r dumpDate`).

```{r, echo=FALSE, message=FALSE}
actualVisits <- intersect(scheduledVisits$Visit, visitInfo$EVENT_ID)
toplot <- merge(
    visitInfo,
    patientData[, "ENROLL_CAT", drop=F],
    by.x="PATNO", by.y=0
)
toplot <- table(toplot[,c("EVENT_ID", "ENROLL_CAT")])
class(toplot) <- "matrix"
toplot <- data.frame(toplot[actualVisits,])
toplot <- cbind(scheduledVisits[actualVisits,], toplot)
kable(toplot, row.names=F)
```

The dates of the different visits for each patient are available in
the `visitInfo` object.
The graphics below
shows an overview of the scheduled visit date for all the PPMI patients
which are currently enrolled.

```{r, echo=FALSE, message=FALSE}
toplot <- visitInfo[which(visitInfo$EVENT_ID %in% rownames(scheduledVisits)),]
toplot <- merge(
    toplot,
    patientData[,"ENROLL_CAT", drop=F],
    by.x="PATNO", by.y=0
)
toplot <- toplot[which(!is.na(toplot$ENROLL_CAT)),]
bp <- boxplot(
    INFODT~EVENT_ID, data=toplot,
    col="grey", outline=FALSE,
    ylim=range(toplot$INFODT, na.rm=T),
    ylab="Date",
    main="Visit date"
)
coln <- sort(as.character(unique(toplot$ENROLL_CAT)))
colLeg <- brewer.pal(n=length(coln), name="Set3")
colLeg <- paste0(colLeg, "AA")
names(colLeg) <- coln
beeswarm(
    formula=INFODT~EVENT_ID, data=toplot,
    add=TRUE,
    pwbg=colLeg[toplot$ENROLL_CAT],
    col="#00000080",
    pch=21,
    corral="density",
    corralWidth=0.7,
    priority="random"
)
rebox(bp, col="black", lwd=c(1.5,3,1.5))
legend(
    "bottomright",
    names(colLeg),
    pch=21,
    col="#00000080",
    pt.bg=colLeg,
    ncol=2,
    bg="transparent"
)
```

## Variables

Among all availables variables,  `r ncol(visitData)` were assessed
at different visit depending on the variable.
These data are available in the `visitData` object.

```{r}
dim(visitData)
visitData[1:5, 1:5]
```

# SNP data

Genotyping of several patients has been done using
Immunochips and NeuroX chips. Data preprocessing is described in
*PPMI\_Methods\_Immunochip\_Genotyping\_108\_121613.pdf* and
*PPMI\_Methods\_NeuroX\_Genotyping\_107\_121613.pdf*
available on the PPMI website. 

These data are available in the `snpData` object.
Basic information about SNPs is available in the `snpInfo` object.

```{r}
print(snpData)
as(snpData[1:5, 1:5], "character")
head(snpInfo)
```

# Preparing data for analysis

## Get cohorts

The function `getCohort` is used to get PATNO (patient IDs) of a cohort of
interest. The function `listCohorts` can be used to get all the available
cohort. Cohorts are defined according to subject enrollment. However among
enrolled subjects some have been withdrawn in the course of the study. These
subjects can be excluded using the `enrolledOnly` parameter.

```{r}
listCohorts()
pd <- getCohort(cohort="PD", enrolledOnly=TRUE)
head(pd)
```

## Extract variables

Variables are extracted for a list of subjects using the `extractVariables`
function.

### Patient variables

```{r}
head(findVarByGroup("Patient"))
pd.d <- extractVariables(
    patients=pd,
    variables=c(
        "ENROLL_CAT", "SimpleGender", "ENROLL_AGE"
    )
)
head(pd.d)
```

### Visit variables

For visit variables the ID of the visit under focus must be provided using
the `event` parameter:

```{r}
head(scheduledVisits)
pd.d <- extractVariables(
    patients=pd,
    variables=c(
        "ENROLL_CAT", "SimpleGender", "ENROLL_AGE",
        "UPDRS1", "UPDRS2", "UPDRS3", "UPDRS"
    ),
    events="BL"
)
head(pd.d)
```

If several visits are requested the function returns a list of data.frames.
One for all the patients variables and one for each visit variable:

```{r}
pd.d <- extractVariables(
    patients=pd,
    variables=c(
        "ENROLL_CAT", "SimpleGender", "ENROLL_AGE",
        "UPDRS1", "UPDRS2", "UPDRS3", "UPDRS"
    ),
    events=c("BL", "V04", "V06")
)
lapply(pd.d, head)
```

This list can be easily stacked in one data.frame by using the following
command:

```{r}
stacked <- do.call(cbind, pd.d)
head(stacked)
```

### SNPs

For SNP data, the ID of the SNPs to be extracted should be
provided using the snps argument.

```{r}
snps <- c("rs356165", "rs3822086", "rs3775424", "rs3857059", "rs2736990")
pd.d <- extractVariables(
    patients=pd,
    variables=c(
        "ENROLL_CAT", "SimpleGender", "ENROLL_AGE",
        "UPDRS1", "UPDRS2", "UPDRS3", "UPDRS"
    ),
    snps=c("rs356165", "rs3822086", "rs3775424", "rs3857059", "rs2736990"),
    transcode=T,
    events=c("BL", "V04", "V06")
)
head(pd.d$Patient)
```

The collected data are bound to the Patient data.frame of the
results.

Information about SNP can be retrieved using the `getSnpInfo`
function.

```{r}
getSnpInfo(snps)
attr(getSnpInfo(snps), "Not found")
```

# Analyses

## SNP analyses with plink

The function `plinkRegression` calls
[plink](http://pngu.mgh.harvard.edu/~purcell/plink/)
to run logistic or
linear regression analysis. In the example below, PD are compared
to HC. Look at `?plinkRegression` to get further information.

```{r}
patients <- c(getCohort("PD"), getCohort("HC"))
phenoTable <- extractVariables(
    patients=patients,
    variables=c("ENROLL_CAT", "ENROLL_AGE", "SimpleGender")
)
phenoTable$ENROLL_CAT <- factor(phenoTable$ENROLL_CAT)
snpSets <- list(
    SNCA=c(
        "rs356165", "NeuroX_rs356165",
        "NeuroX_rs34674090", "NeuroX_rs17180453", 
        "NeuroX_rs35541731", "NeuroX_chr4:90661018",
        "rs3822086", "rs3775424", 
        "rs3857057", "rs36119871", "rs4088093",
        "rs189596", "rs4088094", 
        "NeuroX_rs4088094", "NeuroX_rs28393675",
        "rs3857059", "NeuroX_rs28613708", 
        "NeuroX_rs77490515", "rs2736990",
        "exm-rs2736990", "NeuroX_dbSNP_rs112645122", 
        "NeuroX_rs356198", "NeuroX_rs3910105",
        "NeuroX_rs356197", "NeuroX_rs356196", 
        "NeuroX_rs3910104", "NeuroX_rs10433953", "NeuroX_rs72503734", 
        "NeuroX_chr4:90700236", "NeuroX_rs2572319", "NeuroX_rs10018362", 
        "NeuroX_rs10002435", "NeuroX_rs11097234", "NeuroX_chr4:90715805", 
        "NeuroX_rs6848726", "NeuroX_rs2119787", "NeuroX_rs78679014", 
        "NeuroX_rs35479630", "NeuroX_rs2619343", "NeuroX_rs2737017", 
        "NeuroX_rs2737016", "NeuroX_rs2737014",
        "NeuroX_rs7356297", "NeuroX_rs11097238", 
        "NeuroX_rs3796665", "NeuroX_rs2619357", "NeuroX_rs55932807", 
        "NeuroX_rs79392886", "NeuroX_dbSNP_rs113177203", "exm2241445_ver2",
        "NeuroX_rs3756059", "NeuroX_rs1372518",
        "NeuroX_rs1372519", "rs2245801"
    ),
    GBA=c(
        "imm_1_153471318", "rs368060",
        "rs1064651", "imm_1_153472293",
        "imm_1_153472965", "imm_1_153473592",
        "imm_1_153473654", "imm_1_153473664",
        "imm_1_153473983", "imm_1_153474038",
        "imm_1_153474357", "rs381737",
        "rs1064644", "rs381427",
        "rs381418", "rs364897",
        "imm_1_153474807", "imm_1_153475271",
        "imm_1_153475984", "imm_1_153476586",
        "imm_1_153477542", "imm_1_153477730",
        "imm_1_153477961", "imm_1_153479748",
        "imm_1_153480389", "imm_1_153480480",
        "imm_1_153481097"
    )
)
results <- plinkRegression(
    phenoTable=phenoTable,
    snpSets=snpSets,
    mperm=100,
    type="logistic",
    command="~/bin/plink --noweb"
)
print(head(results$snps[order(results$snps$P),]))
print(results$sets)
```

# Session info

```{r, echo=FALSE}
sessionInfo()
```
