---
title: "PPMI: Data summary and export"
author: "Patrice Godard"
date: "`r format(Sys.time(), '%B %d %Y')`"
abstract: "This document describes the data provided in the PPMI package and how they are exported"
output: 
  BiocStyle::html_document: 
    fig_width: 9
    fig_height: 6.5
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message=FALSE}
rm(list=ls())
null <- gc()
try(null <- dev.off(), silent=T)
rm(null)
library(openxlsx)
library(knitr)
library(RColorBrewer)
library(PPMI)
dumpDate <- as.Date(packageDescription("PPMI")$Date)
xf <- paste0("PPMI-Export-", dumpDate, ".xlsx")
#######
beeswarm <- function(
    formula,
    data,
    corral="density",
    corralWidth=0.8,
    do.plot=TRUE,
    ...
){
    if(corral!="density"){
        toRet <- do.call(beeswarm::beeswarm, c(
            list(
                formula=formula,
                data=data,
                do.plot=F,
                corral=corral,
                corralWidth=corralWidth
            ),
            list(...)
        ))
    }else{
        toRet <- do.call(beeswarm::beeswarm, c(
            list(
                formula=formula,
                data=data,
                do.plot=F
            ),
            list(...)
        ))
        toRet <- do.call(rbind, by(
            toRet,
            toRet$x.orig,
            function(d){
                cpos <- round(mean(d$x, na.rm=T))
                xos <- d$x - cpos
                div <- max(abs(xos), na.rm=T)/(corralWidth/2)
                d$x <- cpos + (xos/div)
                return(d)
            }
        ))
    }
    if(do.plot){
        points(
            toRet$x, toRet$y,
            pch=toRet$pch,
            col=toRet$col, bg=toRet$bg
        )
    }
    invisible(toRet)
}
#######
rebox <- function(bp, boxwex=0.8, ...){
    sp <- boxwex/2
    for(i in 1:ncol(bp$stats)){
        y <- bp$stats[2:4, i]
        segments(
            x0=rep(i-sp, length(y)),
            x1=rep(i+sp, length(y)),
            y0=y,
            y1=y,
            ...
        )
    }
}
```


# Introduction

## Background

In the field of Parkinson’s disease (PD) therapeutics, the ultimate goal is to
develop disease-modifying treatments that slow, prevent or reverse the
underlying disease process. Validated biomarkers of disease progression would
dramatically accelerate PD therapeutics research.
Current progression biomarkers, however, are not optimal and are not
fully validated.
(source: [PPMI website](http://www.ppmi-info.org/about-ppmi/))

## Rationale for PPMI

PPMI (Parkinson's Progression Markers Initiative)
is an observational clinical study to verify progression markers in
Parkinson’s disease. PPMI has emerged as a model for following multiple
cohorts of significant interest and is being conducted at a network of
clinical sites around the world. The study is designed to establish a
comprehensive set of clinical, imaging and biosample data that will be used
to define biomarkers of PD progression.
Once these biomarkers are defined, they can be used in therapeutic studies,
which is the ultimate goal.
(source: [PPMI website](http://www.ppmi-info.org/about-ppmi/))

PPMI will follow standardized data acquisition protocols to ensure that
tests and assessments conducted at multiple sites and across multiple cohorts
can be pooled in centralized databases and repositories.
The clinical, imaging and biologic data will be easily accessible
to researchers in real time through
the [PPMI website](http://www.ppmi-info.org/about-ppmi/).
The biological samples collected throughout the course of PPMI will
be stored in a central repository that will be accessible to any scientist
with promising biomarker leads for the purposes of verifying initial
results and assessing correlations to clinical outcomes and other biomarkers.
(source: [PPMI website](http://www.ppmi-info.org/about-ppmi/))

## Data source

Data were dumped on
`r format(dumpDate, "%A %B %d %Y")` from the
[PPMI repository](https://www.ppmi-info.org/access-data-specimens/download-data/). They
were then preprocessed and integrated and derived variables were computed as
recommended by the PPMI.

In this document we describe the data which were preprocessed and how they
were exported in the [`r xf`](`r xf`) file.

# Patient cohorts

The PPMI is focused on
`r length(setdiff(unique(patientData$ENROLL_CAT), NA))` patient cohorts:

```{r, echo=FALSE, message=FALSE}
toplot <- unique(patientData[,c("ENROLL_CAT", "APPRDX")])
toplot <- toplot[which(!is.na(toplot$ENROLL_CAT)),]
toAdd <- table(patientData[,c("ENROLL_CAT", "ENROLL_STATUS")])
class(toAdd) <- "matrix"
toAdd <- data.frame(toAdd[,c("Enrolled", "Withdrew")])
toplot <- merge(toplot, toAdd, by.x="ENROLL_CAT", by.y=0)
colnames(toplot) <- c(
    "Enrollment Category", "Appropriate Diagnosis",
    "Enrolled and not withdrew", "Enrolled and withdrew"
)
kable(toplot, row.names=F)
```

These cohorts are described in this
[PPMI webpage](http://www.ppmi-info.org/study-design/study-cohorts/).

# Variable documentation

## Groups of variables

Variables have been classified in groups and sub-groups. In total, there are
`r nrow(varDoc)` variables belonging to
`r nrow(unique(varDoc[,c("Group", "Sub-group")]))` sub-groups in
`r length(unique(varDoc$Group))` groups.
This information is summarized in the table below.

```{r, echo=FALSE, message=FALSE}
kable(variableGroupSummary(), row.names=F)
```

## Classes and source of variables

Variables can be of different classes as shown in the following table.
*Character* variables correspond to comments precising patient status or
treatment.

```{r, echo=FALSE, message=FALSE}
toplot <- as.data.frame(table(varDoc$Class))
colnames(toplot) <- c("Class", "Number of variables")
kable(toplot, row.names=F)
```

Variables have been either directly extracted from files provided by
the PPMI or derived according to PPMI recommendations:

```{r, echo=FALSE, message=FALSE}
toplot <- as.data.frame(table(varDoc$Source))
colnames(toplot) <- c("Source", "Number of variables")
kable(toplot, row.names=F)
```

## Export

```{r, echo=FALSE, message=FALSE}
xs.varDoc <- "Variable-Documentation"
```

Documentation of variables have been exported in the *`r xs.varDoc`*
spreadsheet of the [`r xf`](`r xf`) file.

# Patient variables

```{r, echo=FALSE, message=FALSE}
xs.patientData <- "Patient-Data"
```

Among all availables variables, `r ncol(patientData)` were assessed only once,
during the recruitment. They belong to the group of variable named *Patient*.
These variables have been exported in the *`r xs.patientData`*
spreadsheet of the [`r xf`](`r xf`) file.

```{r, echo=FALSE, message=FALSE}
varDocExport.p <- data.frame(
    Variable=colnames(patientData),
    varDoc[colnames(patientData),],
    stringsAsFactors=F,
    check.names=F
)
for(cohort in names(listCohorts())){
    patno <- getCohort(cohort)
    subPat <- patientData[patno, ]
    varDocExport.p[,cohort] <- apply(
        subPat,
        2,
        function(x) sum(!is.na(x))
    )
}
```

Numbers of available data for each cohort have been exported in
the *`r xs.varDoc`* spreadsheet of the [`r xf`](`r xf`) file.

# Follow-up variables

## Visits

Patients condition is assessed during different visits. The possible visits are
described in the table below.

```{r, echo=FALSE, message=FALSE}
kable(visitDoc, row.names=F)
```

Among these visits, some are scheduled from the begining of the study. The
table below shows the number of patients which were followed-up at each visit
when the data were downloaded (`r dumpDate`).

```{r, echo=FALSE, message=FALSE}
actualVisits <- intersect(scheduledVisits$Visit, visitInfo$EVENT_ID)
toplot <- merge(
    visitInfo,
    patientData[, "ENROLL_CAT", drop=F],
    by.x="PATNO", by.y=0
)
toplot <- table(toplot[,c("EVENT_ID", "ENROLL_CAT")])
class(toplot) <- "matrix"
toplot <- data.frame(toplot[actualVisits,])
toplot <- cbind(scheduledVisits[actualVisits,], toplot)
kable(toplot, row.names=F)
```

```{r, echo=FALSE, message=FALSE}
xs.visitInfo <- "Visit-Date"
```

The dates of the different visits for each patient have been exported in
the *`r xs.visitInfo`* spreadsheet of the [`r xf`](`r xf`) file.
The graphics below
shows an overview of the scheduled visit date for all the PPMI patients
which are currently enrolled.

```{r, echo=FALSE, message=FALSE}
toplot <- visitInfo[which(visitInfo$EVENT_ID %in% rownames(scheduledVisits)),]
toplot <- merge(
    toplot,
    patientData[,"ENROLL_CAT", drop=F],
    by.x="PATNO", by.y=0
)
toplot <- toplot[which(!is.na(toplot$ENROLL_CAT)),]
bp <- boxplot(
    INFODT~EVENT_ID, data=toplot,
    col="grey", outline=FALSE,
    ylim=range(toplot$INFODT, na.rm=T),
    ylab="Date",
    main="Visit date"
)
coln <- sort(as.character(unique(toplot$ENROLL_CAT)))
colLeg <- brewer.pal(n=length(coln), name="Set3")
colLeg <- paste0(colLeg, "AA")
names(colLeg) <- coln
beeswarm(
    formula=INFODT~EVENT_ID, data=toplot,
    add=TRUE,
    pwbg=colLeg[toplot$ENROLL_CAT],
    col="#00000080",
    pch=21,
    corral="density",
    corralWidth=0.7,
    priority="random"
)
rebox(bp, col="black", lwd=c(1.5,3,1.5))
legend(
    "bottomright",
    names(colLeg),
    pch=21,
    col="#00000080",
    pt.bg=colLeg,
    ncol=2,
    bg="transparent"
)
```

## Variables

```{r, echo=FALSE, message=FALSE}
xs.visitData <- "Visit-Data"
```

Among all availables variables,  `r ncol(visitData)` were assessed
at different visit depending on the variable.
These variables have been exported in the *`r xs.visitData`*
spreadsheet of the [`r xf`](`r xf`) file.

```{r, echo=FALSE, message=FALSE}
varDocExport.e <- data.frame(
    Variable=colnames(visitData),
    varDoc[colnames(visitData),],
    stringsAsFactors=F,
    check.names=F
)
for(cohort in names(listCohorts())){
    patno <- getCohort(cohort)
    for(v in scheduledVisits$Visit){
        visCoh <- paste(cohort, v, sep=" - ")
        ids <- paste(patno, v, sep=".")
        subVis <- visitData[ids,]
        varDocExport.e[, visCoh] <- apply(
            subVis,
            2,
            function(x) sum(!is.na(x))
        )
    }
}
```

Numbers of available data for each cohort and each visit have been exported in
the *`r xs.varDoc`* spreadsheet of the [`r xf`](`r xf`) file.

<!-------------------------------------------------->
<!-------------------------------------------------->

```{r, echo=FALSE, message=FALSE}
## Duplicated lower case column names
varDocExport <- merge(varDocExport.p, varDocExport.e, all=T)
toRename <- which(duplicated(tolower(varDocExport$Variable)))
toRename <- unique(tolower(varDocExport$Variable)[toRename])
renamed <- c()
for(dvn in toRename){
  ntoadd <- varDocExport$Variable[which(
    tolower(varDocExport$Variable)==dvn
  )]
  toadd <- paste0(ntoadd, " (", 1:length(ntoadd), ")")
  names(toadd) <- ntoadd
  renamed <- c(renamed, toadd)
}
toadd <- setdiff(varDocExport$Variable, names(renamed))
names(toadd) <- toadd
renamed <- c(renamed, toadd)
varDocExport$Variable[match(names(renamed), varDocExport$Variable)] <- renamed
## Patient data
patientData <- PPMI::patientData
colnames(patientData) <- renamed[colnames(patientData)]
visitData <- PPMI::visitData
colnames(visitData) <- renamed[colnames(visitData)]
```


```{r, echo=FALSE, message=FALSE}
## Final xlsx export
varDocExport <- varDocExport[
    order(paste(varDocExport$Group, varDocExport$`Sub-group`, sep="-->")),
]
wb <- createWorkbook()
##
addWorksheet(wb, xs.varDoc)
writeDataTable(
    wb, xs.varDoc, varDocExport, rowNames=FALSE, colNames=TRUE,
    startRow=1, startCol=1, tableStyle = "TableStyleLight9"
)
setColWidths(wb, xs.varDoc, cols = 1:(ncol(varDocExport)+1), widths = "auto")
##
addWorksheet(wb, xs.patientData)
writeDataTable(
    wb, xs.patientData, patientData, rowNames=TRUE, colNames=TRUE,
    startRow=1, startCol=1, tableStyle = "TableStyleLight9"
)
setColWidths(wb, xs.patientData, cols = 1:(ncol(patientData)+1), widths = "auto")
##
addWorksheet(wb, xs.visitInfo)
writeDataTable(
    wb, xs.visitInfo, visitInfo, rowNames=TRUE, colNames=TRUE,
    startRow=1, startCol=1, tableStyle = "TableStyleLight9"
)
setColWidths(wb, xs.visitInfo, cols = 1:(ncol(visitInfo)+1), widths = "auto")
##
addWorksheet(wb, xs.visitData)
writeDataTable(
    wb, xs.visitData, visitData, rowNames=TRUE, colNames=TRUE,
    startRow=1, startCol=1, tableStyle = "TableStyleLight9"
)
setColWidths(wb, xs.visitData, cols = 1:(ncol(visitData)+1), widths = "auto")
##
saveWorkbook(wb, xf, TRUE)
```


